Decentralized state machine using Nakamoto (probabilistic) consensus\\\medskip Research Project 2022

### 2022-11-01 article/article.tex
typos

In this article, we look at state machines as a way to hold in multiple servers copies of the same record. That record, in my implementation, is a standard
key-value data structure (\href{https://go.dev/blog/maps}{a Go map}).

@@

Many classical consensus algorithms exists, the most famous one being Paxos \cite{parliament}. In my project, I used the Raft consensus algorithm \cite{understandable} because it has the same performances
as Paxos, and is simpler to understand and implement.

@@

The only parameter than I included in my implementation is \emph{updateSystem}, which is a boolean. If True, if the leader sends heartbeats and a node does not reply, it will be eliminated from the network in the knowledge of all the nodes. This make it possible to
crash a majority of nodes and still have the state machine running and usable on all the surviving nodes.

@@

Contrary to Crash Fault, Byzantine Fault \cite{byzantine} is a type of failure where you consider that the node can fail in any possible way. In practice, it means that the node will stay up and can send malicious requests to
interfere with the functioning of the network.

My implementation of the state machine does not tolerate Byzantine Faults at all. With only one byzantine node, it already endangers the whole system. Indeed, the malicious node can "elect himself" in Raft, and then send heartbeats to all
the node in the network, including the legitimate leader to become the leader de facto. Then, it can ignore requests, reply whatever it wants and write any log it wants in any node it wants.

To tackle this issue, BFT algorithms were designed (such as \cite{pbft}). They include a layer of authentification, but are not necessarily much slower. However, there is a maximum number of malicious nodes that the algorithm can handle safely \cite{partialSynchrony}.
For example, assuming partial synchrony of the nodes, consensus protocols can handle t crash failures with 2t+1 nodes but t byzantine failures with no less than 3t+1 nodes.

@@

Nakamoto consensus was introduced with the bitcoin blockchain \cite{bitcoin}. Nakamoto consensus is probabilistic, which means that there is a probability ε that consensus is not reached.
Of course, ε can be so low that crucial systems can be built relying Nakamoto consensus. Nakamoto consensus is open : nodes don't have to register or get an authorisation to enter the network.
Because nodes don't have to know every other node in the network and be able to communicate with everyone, Nakamoto is supposed to allow for greater scale in the network.

Nakamoto consensus definetly shaked the consensus space in computer science with cryptocurrencies. In a different context, can it be useful as well ?

@@

I adapted the protocol to fit the needs of a decentralized state machine, because snowball is described in the context of building a payment system (because it is the basis of Avalanche, which is that payment system).
Snowball is built by iterations, but I will only focus on Snowball, the last iteration, because it is the most interesting and the one I tested the performances in the end. Slush and Snowflake are interesting to understand
where we are coming from, but no more.

@@

In snowball, each node has a counter of "confidence" which can tip towards one way or another. In my implementation, there is no binary consensus because the only value the system has is the log request.
When receiving an add-log request, the node follows this sequence, in which $logToAdd$($string$) and $logPosition$($int$) are variables from the request (client or other node) and $record$($map[int]string$) and $cnt$($map[int]int$) are variables from the node memory:
\begin{algorithmic}
    \If{$record[logPosition]$ exists}
    \State reply $record[logPosition] == logToAdd$
    \Else
    \State $record[logPosition] \gets logToAdd$
    \State reply $True$
    \State initiateRequest($logToAdd$, $logPosition$)
    \EndIf
\end{algorithmic}

@@

\begin{algorithmic}
    \While{$True$}
    \State $success \gets initiateQuery(req)$
    \State $initiateQuery$ picks a sample in the network and queries it
    \If{$success$}
    \State $cnt[logPosition]++$
    \Else
    \If{$cnt[logPosition] == 1$}
    \State delete($record$, $logPosition$)
    \State $cnt[logPosition] \gets 0$
    \Else
    \State $cnt[logPosition]--$
    \EndIf
    \EndIf
    \If{$cnt[logPosition] \geqslant CounterThreshold$}
    \State $record[logPosition] \gets logToAdd$
    \State return
    \EndIf
    \EndWhile
\end{algorithmic}

3 parameters are used in my implementation :
\begin{itemize}
    \item sampleSize (k) : the number of nodes that are queried each round
    \item majorityThreshold (α) : the number of nodes that have to reply successfully to consider that the query is successfull.
          (with $\alpha \geq \lceil k/2 \rceil$)
    \item counterThreshold (β) : the threshold of successfull requests above which the node stops initiating requests and only replies with its content. (for one position)
\end{itemize}

@@

In order to compare the performances of classical consensus and Nakamoto consensus, we need to be able to know the probability of not reaching consensus ε corresponding to the set
of values (n, k, β). (α is not included because we assume no byzantine nodes in this part).

@@

Then, the probability of reaching a certain state (1) is the sum of the probability of a different state (2) times the probability of going from state 2 to state 1.
That is true if and only if states are transient, which is the case here because c can only grow and with c constant, reqs can only diminish.

@@

The term in the sums is considered 0 if it does not make sense.
First sum represents a step of the resolution of one group of requests, coming from a semi-colored node, with tc (ts) the number of colored (semi colored) nodes in the sample.
Second sum represents a step of the resolution of one group of requests, coming from a colored node.
To do the calculation, we assume that requests from semi-colored nodes resolve faster.
\begin{multline*}
    P(c, s, rc, rs) = \\ \sum_{tv=0}^{k+1}\sum_{ts=0}^{k+1} {n-c-s+tv \choose tv}{s-tv-1+ts \choose ts}{c-ts \choose k-tv-ts}P(c-ts, s-tv+ts, rc-ts, rs+1-tv)
    \\ + \sum_{ts=0}^{k+1} {n-c-s+rs \choose tv}{s-rs+ts \choose ts}{c-ts-1 \choose k-rs-ts}P(c-ts, s-rs+ts, rc+1-ts, 0)
\end{multline*}
\\
With these formulas, we can compute the probability of not reaching consensus : view Appendices
The scripts which were used to compute the graphs in Appendix are available at \href{https://github.com/Nicolascrd/researchProjectConsensus}{github.com/Nicolascrd/researchProjectConsensus}

@@

We demonstrated that it was possible to build a decentralized state machine, that could be based on a Nakamoto consensus algorithm and had interesting properties.
However, more research is required to use this approach at a greater scale.
In particular, it would be interesting to approximate the value of ε $\forall n, k, \beta$. 
Also, a comparison focused on Byzantine Fault Tolerance would be interesting. To do that one would need to use PBFT \cite{pbft} instead of the Raft.

>>>>

In this article, we look at state machines as a way to hold in multiple servers copies of the same record. In my implementation, this record is a standard
key-value data structure (\href{https://go.dev/blog/maps}{a Go map}).

@@

Many classical consensus algorithms exists, the most famous one being Paxos \cite{parliament}. In my project, I used the Raft consensus algorithm \cite{understandable} because it performs as well as Paxos, and is simpler to understand and implement.

@@

The only parameter than I included in my implementation is \emph{updateSystem}, which is a boolean. If $True$, if the leader sends heartbeats and a node does not reply, 
it will be eliminated from the network in the knowledge of all the nodes. This makes it possible to
crash a majority of nodes and still have the state machine running and usable on all the surviving nodes.

@@

Contrary to Crash Fault, Byzantine Fault \cite{byzantine} is a type of failure where you consider that the node can fail in any possible way. In practice, it means that the node will stay up and can send malicious requests to
interfere with network usage.

The Raft consensus algorithm is not BFT. BFT algorithms (such as \cite{pbft}) include a layer of authentification, but are not necessarily much slower. However, there is a maximum number of malicious nodes that the algorithm can handle safely \cite{partialSynchrony}.
For example, assuming partial synchrony of the nodes, consensus protocols can handle $t$ crash failures with $2t+1$ nodes but $t$ byzantine failures with no less than $3t+1$ nodes.

@@

Nakamoto consensus was introduced with the bitcoin blockchain \cite{bitcoin}. Nakamoto consensus is probabilistic, which means that there is a probability ε that consensus is not reached.
Of course, ε can be so low that crucial systems can be built relying on Nakamoto consensus. Nakamoto consensus is open : nodes don't have to register or get an authorisation to enter the network.
Because nodes don't have to know every other node in the network and be able to communicate with everyone, Nakamoto is supposed to allow for greater scale in the network.

Nakamoto consensus was a revolution in the consensus space with cryptocurrencies. How can we adapt the approach to build a state-machine with it ?

@@

I adapted the protocol to fit the needs of a decentralized state machine, because snowball is described in the context of building a payment system (it is the basis of Avalanche, which is that payment system).
Snowball is built by iterations, but I will only focus on the last iteration, because it is the most interesting and the one I tested the performances in the end. 
The other two algorithms, Slush and Snowflake are only interesting to understand the protocol.

@@

In snowball, each node has a counter of "confidence" which can tip towards one way or the other. In my implementation, there is no binary consensus because the only value the system has is the log request.
When receiving an add-log request, the node follows this sequence, in which $logToAdd$(string) and $logPosition$(int) are variables from the request (client or other node) and $record$(map[int]string) and $cnt$(map[int]int) are variables from the node memory:
\begin{algorithmic}
    \If{$record[logPosition]$ exists}
    \State reply $record[logPosition] == logToAdd$
    \Else
    \State $record[logPosition] \gets logToAdd$
    \State reply $True$
    \State initiateRequest($logToAdd$, $logPosition$)
    \EndIf
\end{algorithmic}

@@

\begin{algorithmic}
    \While{$True$}
    \State $success \gets initiateQuery(req)$
    \State ($initiateQuery$ picks a sample in the network and queries it)
    \If{$success$}
    \State $cnt[logPosition]++$
    \Else
    \If{$cnt[logPosition] == 1$}
    \State delete($record$, $logPosition$)
    \State $cnt[logPosition] \gets 0$
    \Else
    \State $cnt[logPosition]--$
    \EndIf
    \EndIf
    \If{$cnt[logPosition] \geqslant CounterThreshold$}
    \State $record[logPosition] \gets logToAdd$
    \State return
    \EndIf
    \EndWhile
\end{algorithmic}

3 parameters are used in my implementation :
\begin{itemize}
    \item sampleSize (k) : the number of nodes that are queried each round
    \item majorityThreshold (α) : the number of nodes that have to reply successfully to consider that the query is successfull.
          (with $\alpha \geq \lceil k/2 \rceil$)
    \item CounterThreshold (β) : the threshold of successfull requests above which the node stops initiating requests and only replies with its content. (for one position)
\end{itemize}

@@

In order to compare the performances of classical consensus and Nakamoto consensus, we need to be able to know the probability of not reaching consensus ε corresponding to the set
of values (n, k, β). (α is not included because we assume no byzantine nodes in this section).

@@

Then, the probability of reaching a certain state ($S1$) is the sum of the probability of a different state ($S2$) times the probability of going from state $S2$ to state $S1$.
That is true if and only if states are transient, which is the case here because c can only grow and with c constant, reqs can only diminish.

@@

The term in the sums is considered 0 if it does not make sense.
First sum represents a step of the resolution of one group of requests, coming from a semi-colored node, with tc (ts) the number of colored (semi colored) nodes in the sample.
Second sum represents a step of the resolution of one group of requests, coming from a colored node.
To do the calculation, we assume that requests from semi-colored nodes resolve faster.
\begin{multline*}
    P(c, s, rc, rs) = \\ \sum_{tv=0}^{k+1}\sum_{ts=0}^{k+1} {n-c-s+tv \choose tv}{s-tv-1+ts \choose ts}{c-ts \choose k-tv-ts}P(c-ts, s-tv+ts, rc-ts, rs+1-tv)
    \\ + \sum_{ts=0}^{k+1} {n-c-s+rs \choose tv}{s-rs+ts \choose ts}{c-ts-1 \choose k-rs-ts}P(c-ts, s-rs+ts, rc+1-ts, 0)
\end{multline*}
\\
With these formulas, we can compute the probability of not reaching consensus : view Appendices.
The scripts which were used to compute the graphs in Appendix are available at \href{https://github.com/Nicolascrd/researchProjectConsensus}{github.com/Nicolascrd/researchProjectConsensus}

@@

We demonstrated that it was possible to build a decentralized state machine, that could be based on a Nakamoto consensus algorithm and had interesting properties. 
In normal running state, the number of requests to reach consensus is a function of $k * \beta$ whereas in Raft, the leader must handle $n-1$ request on its own. 
Raft has one consensus for the whole system, whereas in Snowball to each value correspond a different consensus. For most usage, classical consensus is probably the way to go.
It would be useful to lead some more research on the topic. 
In particular, it would be interesting to approximate the value of ε $\forall n, k, \beta$. 
Also, a comparison focused on Byzantine Fault Tolerance would also be interesting. To do that one would need to use PBFT \cite{pbft} instead of the Raft.
